Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-02-21 paper.qmd
lot of additions to the grammar of graphics section

## Types of Uncertainty Visualisation Papers
There are two broad categories of uncertainty visualisation experiments, depending on what they believe the "problem" in uncertainty visualisation to be.

The first type of paper believes uncertainty to be fundamentally differently to other variables due to psychological elements of uncertainty, such as risk aversion. These papers seek to investigate uncertainty visualisaiton by assessing which visualisation methods facilitate these "risk" related biases and check that perceptual tasks are not impacted by the fact that "uncertainty" is what is being visualised. At least this is often how the motivation is described. Due to the fact that uncertainty is poorly defined and the psychological effects of uncertainty is never justifiably connected to the scientific communication of it, as was discussed in ssections 3 and 4, these papers are all over the place in terms of methodology. Since it is unclear exactly *what* they are testing for, since it is unclear *how* uncertainty information is suppoesdly different from any other information, these papers are poorly justified and attempt to test for the secondary effects of a psychological outcome that has actually never been justified or shown to exist.

The second type of paper perceives uncertainty to be the same as any other variable, and therefore the issue facing uncertainty visualisation is actually a dimensionality problem. These experiments evaluate different ways to include even *more* information than would be put in a typical visualisation. Some of these methods, such as a bivariate pallet or glyph map, create graphics that are overloaded with information and offer incorrect or meaningless takeaways at first glance. This is not to say *every* visualisation method that sees it as a dimensionality problem is bad. Clever methods of combining several types of information, such as value supressing uncertainty pallets (VSUPs), can better help us understand how our brains process information in a graphic. However this is because visualisations that simply seek to display as much information as possible fail to understand visualisation are a tool to summarise information using visual features, and once it fails to summarise information it is no longer a good visualisation. Failling to understand this key aspect of visualsiation leads to graphics that *contain* the correct amount of information, but that volume of information cannot be readily accessed by the viewer.

In addition to the way the papers *see* the uncertainty visualisation problem, these papers can also be organised according to the *goal* of the experiment. Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation. Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data. This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study. Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work. Until that later work is done, it is often difficult to accept one particular representation. While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.

@@

## Uncertainty Visualisation and the Grammar of Graphics
The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty.

- Since there are so many papers that treat uncertainty as a signal, not a noise, issues in the uncertainty visualisation literature also help to understand some difficulties in the information visualisation literature as large.

>>>>

## Types of Uncertainty Visualisation Papers
Uncertainty visualisation papers can be organised according to the *goal* of the experiment. Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation. Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data. This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study. Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work. Until that later work is done, it is often difficult to accept one particular representation. While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.

@@

## Uncertainty Visualisation and the Grammar of Graphics
The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty. This problem seems to start at the lack of the mathematical definition of uncertainty which then bleeds into a lack of a definition for an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a large number of uncertainty visualisation experiments. How can a chart omit uncertainty according to the grammar of graphics, but be used to describe it in several evaluation studies? The answer to this question is related to the definition problem discussed earlier, but to understand how, we will need to go through the grammar of graphics, and its absence in the uncertainty visualisation literature.

When creating a graphic there are several tasks that must be completed in a specific order, regardless of whether or not we are working with a uncertainty visualisation or not. @Leland2005 depicts these steps as a pipeline, shown in @fig-pipeline. This pipeline shows every step from data observation to rendering of a graphic, but we are going to focus on the tail end of the pipeline specifically on statistics, geometry, coordinates, and aesthetics.

![The grammar of graphics data analysis pipeline](grammarofgraphicspipeline.png){#fig-pipeline}

In order to visually communicate uncertainty we must:  
1) Mathematically quantify uncertainty in some way. 
2) Visually express this quantification using some geometry, axis, aesthetic, and rendering.
3) Interpret the visual expression to extract the information using perceptual tasks.

Each of these steps have a large amount of research independent of the "uncertainty visualisation" subfield. There are numerous ways to depict uncertainty mathematically. Examples of mathematically defined objects that are used to express uncertainty inlude: confidence intervals, sampling distributions, bias, variance, prediction error, value ranges, quantiles etc. These definitions all exist in mathematics independent of the field of visualisation. Once we have some quantification of uncertainty, we need to express it visually. There are numerous options in combinations of geometry, axis, and aesthetics and modern graphical software will allow most of these combinations. There is also a large amount of research in information visualisation and psychology into how the choice of visual expression impacts a persons ability or ease with which they extract relevant information. In this sense, once we have a mathematical expression of uncertainty, the visualisation of that uncertainty is identical to the visualisation process of any other variable, or at least this is how it is treated in the uncertainty visualisation experiments.

This pipeline leads to two obvious facts that are *well* established in mathematics and visualisation. First, a graphic that depicts one statistic does not necessary contain information about a different, possibly irrelevant, statistic. For example, if you compare two graphics and one depicts the mean and the other depicts a median and you ask viewers to extract the median value, the graphic that *actually* depicts that statistic will perform better. People cannot percieve information that is not depicted. This is not necessarily a feature of visualisation, this is simply a fact of mathematics. Additionally, if information that can be used to calculate a statistic, but not the statistic itself, is shown, people will need to use a heuristic to extract that information which will decrease the precision with which that information is communicated. Second, there is a hierarchy to perceptual tasks where extracting visual information in some forms is easier than others. The hierarchy established 40 years ago by @Cleveland1984 is: 
1) position along a common scale
2) position along a non-aligned scale
3) length
4) angle/slope
5) area
6) volume
7) colour
This hierarchy has had additions and adjustments over the years. For example, this hierarchy is a general rule, but the hierarchy can change from individual to individual. Additionally, there are other graphical rules to consider such as guestalt heuristics, broader methods of extraction, and attention principles. These establishe visualisation concepts allow us to anticipate the ease with which certain pieces of information will be extracted from a plot. A bizarre feature of uncertainty visualisation literature is that many of the evaluation studies seem to be written as though these mathematical facts and established visualisaiton rules do not exist.

The papers that seem to be the most unaware of these established rules in visualisation are the plot comparison papers. The most common type of uncertainty visualisation evaluation study is a "comparison of plots". The papers that show several uncertainty visualisations and ask participants to extract statistics. In these papers will typically ask separate questions about the signal (e.g. what is the mean of this distribution) and the uncertainty (e.g. what is the variance of this distribution). Then the accuracy, speed, or confidence of the participants responses are compared across the plots. The overwhelming majority of these papers depict different mathematical expressions of both the signal and the uncertainty. In 100% of these papers, the plot that has mapped the statistic in question to some visual feature will outperform those that don't. When the statistics used to depict uncertainty *are* the same (which is rare) 100% of those papers will follow the hierarchy of perceptual tasks that has already been established in the visualisation literature. If the experiment participant needs to extract a probability and one graphic uses position along a common scale, while the other uses colour or literally any other perceptual task, the position depiction will always outperform the other. Strangely, none of these papers comment that their study seems to simply follow established perceptual task rules. Instead they opt to comment on how a "bar chat outperforms a mean with confidence intervals" not seeming to connect that it was only on questions related to the median. This makes it clear we have a serious problem with the classification of "uncertainty visualisation". The visualisations follow identical rules to those established already in the visualisation literature, and experiments that view uncertianty visualisation as somehow special or different from normal visualisation are simply repeating existing established research in information visualisation. This stems from a class of "uncertainty visualisation" that seems to have no rules for what belongs in the group, and no consideration for the mathematics that is actually depicted in these plots. Uncertainty visualisation experiments do not work to build upon existing visualisation literature or differentiate uncertainty visualisaiton as a new field, but seem to exist with shocking indifference to the field it belongs to.

This brings us to a second problem in these papers. Since there is no rules for what is or is not an uncertainty visualisation, graphics that are compared in this sub-dicipline are often *wildly* different. If we were to compare two different visualisations, we would like to change as few features as possible so we can better understand the cause of the difference in perception. If several graphics are compared that are different statistically, geometrically, and aesthetically, there is *no way to combine the results to establish generic principles for the field*. Uncertainty visualisaiton papers only seem to require that the plots depict "uncertainty" or typically belong to the class of "uncertainty visualisation", some of which are simply invented by the author. The complete lack of formality is easy to see if we actually plot the visualisations compared across evaluation experiments according to the grammar of graphics *in the figure below*:

**INCLUDE A GRAPHIC THAT SHOWS WHERE EACH TYPE OF UNCERTAINTY VISUALISATION SITS IN THE GRAMMAR OF GRAPHICS AND HOW DIFFERENT THEY ARE ACROSS MULTIPLE STUDIES**

This lack of formality has severly cost the field. This lack of formalisation is not unique to uncertainty visualisation and the wider field of information visualisation suffers form similar problems [@Kinkeldey2014]. Several uncertainty visualisation literature reviews have commented on the immense difficulty of combining results across experiments, but none have been able to pinpoint the reason. @Kinkeldey2014 suggested a task based taxonomy that organised the plots by the question or purpose of the visualisation, @Spiegelhalter2017 simply ignored the existence of uncertainty visualisation as a subfield and just suggested we follow established rules within uncertainty visualisation (which was actually good advice), @Hullman2016 commented on the large amount of noise generated from uncertainty visualisation evaluation methods that led to a difficulty in combining results, @Griethe2006 did not meaningfully combine the work and instead listed findings in the field, but also did not comment on the lack of generalisable results. While we largely agree with the difficulties of these authors, but instead of a task based taxonomy or better questions, we believe uncertainty visualisation, as it is now, should not exist. Any generalisable contribution made by these papers is already an established result in information visualisaiton, and therefore is not specific to "uncertainty visualisation". Any "uncertainty visualisation" specific result is not generalisable beyond the experimental conditions because of the poor formalisation in the definitions of uncertainty and uncertainty visualisations.

Exactly *why* uncertainty visualisation is different from typical visualisation is rarely discussed, however there does seem to be some subtextual arguments put forth in the motivation of these expermients. These motivations often discuss at least one of two reasons uncertainty visualisation authors seem to think uncertainty visualisation should be treated differently to normal visualisation.
1) Uncertainty is fundamentally differently to other variables due to psychological heuristics involved in the interpretation of uncertainty, such as risk aversion.
2) Uncertainty is not of interest in of itself and is typically layered on at the end of a the visualisation pipeline, so uncertainty visualisation is the field of adding error information into already established graphics.

Papers that are justified using the first type of motivation rarely explain exactly how the psychological apsects of uncertainty impct the participants in a way that inhibits their ability to read information from a graphic. Since uncertainty is poorly defined and the psychological effects of uncertainty is never justifiably connected to the scientific communication of it, as was discussed in ssections 3 and 4, it is unclear exactly *what* these papers are testing for. It is almost as though they *feel* that uncertainty visualisations *should* be different, but do not consider *how* uncertainty information is supposedly different beyond decision making criteria such as *risk avoidance* which is not something science communicators have any business trying to control, as was discussed in previous sections. For these reasons, experiments that test for secondary factors (such as influence on decision making) are confounding far too many variables to establish any generalisiable take away, and experiments that ask participants to extract values fail to notice there is no longer a reason for "uncertainty" to be any different to any other type of visualisation.

Papers that are justified using the second type of motivation clearly fit within the sub-field of high dimensional visualisation, but also rarely comment on the connection. These experiments evaluate different ways to include even *more* information than would be put in a typical visualisation. Some of these methods, such as a bivariate pallet or glyph map, create graphics that are overloaded with information and offer incorrect or meaningless takeaways at first glance. This is not to say *every* visualisation method that sees it as a dimensionality problem is bad. Clever methods of combining several types of information, such as value supressing uncertainty pallets (VSUPs), can better help us understand how our brains process information in a graphic.  For example, the VSUP allows us to test is colour value and saturation can be combined such that we mentally combine signal and noise together as one variable. Since the take aways from the papers are not usually related to uncertainty specifically, and often related to interesting perceptual tasks, there is also no justification for why this work is "uncertainty visualisation" work and not simply perceptual task studies that can be used by the rest of the information visualisation community.

These two perceptualisations of *why* uncertainty visualisation is different from normal visualisation is also what leads to differing opinions on what is an uncertainty visualisation. For example, someone who percieves of uncertainty visualisation as a dimensionality problem would see a pie chart as a normal graphic, and a pie chart with blurred boundaries as an uncertainty visualisation, while someone who sees uncertainty visualisation as psychologically differentwould see a pie chart as an uncertainty graphic so long as it depicts probabilities. This may lead you to believe that those who believe uncertainty is unique due to its psychological properties believe uncertainty should be defined at the mathematical stage, while those that believe it is an addition to an established visualisation think uncertainty is defined in a graphic at the aesthetic stage. This is not the case. As shown by the *grammar of graphics fig* there is no single person or experiment that is consistent in any one belief. This difficulty and lack of consistency comes from the desire to discuss "uncertainty visualisation" as though it is a single term and not the combination of "uncertainty" and "visualisation". If we split the terms and look at the quantification and visualisation of uncertainty separately, the results of many uncertainty visualisation papers become self evident.

This is not to say any paper that discusses uncertainty visualisation is completely pointless because the results already exist. Perceptual task experiments that identify

and fruitless endeavor. Rather, we want to established that the current literature seems to be disconnected from the actual motivations of uncertainty visualisation.
